Visualizing Contingency Tables

### 2006-09-03 trunk/vcd/paper/Handbook/hsc.Rnw
highlighting_fill now also accepts a color vector.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@552 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Indexing done.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@555 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\usepackage{Sweave,hyperref,color,graphicx} 
\usepackage[bottom]{footmisc}% places footnotes at page bottom

@@

@@

Categorical data analysis is typically based on two- or
higher-dimensional contingency tables, cross-tabulating the
co-occurences of levels of nominal and/or ordinal data. 
In order to explain these, statisticians
typically look for (conditional) independence structures using common
methods such as independence tests and
log-linear models. One idea of visualization techniques is to
use the human visual system to detect structures in the data that
possibly are not obvious from solely numeric output (e.g., test
statistics). Whether the
task is purely exploratory or model based, techniques such as mosaic, sieve, and
association plots offer good support for visualization. Especially
mosaic and sieve plots have been extended over the last two decades, 
and implementations exist in many statistical environments.

All three graphical methods visualize aspects of (possibly higher-dimensional)
contingency tables. A \emph{mosaic plot} \cite{vcd:Hartigan+Kleiner:1984} is basically 
an area-proportional visualization of (typically, observed) frequencies, composed
of tiles (corresponding to the cells) created by recursive
vertical and horizontal splits of a rectangle. Thus, the area of each tile
is proportional to the corresponding cell entry \emph{given} the
dimensions of previous splits. 
\emph{Sieve plots} \cite{vcd:riedwyl+schuepbach:1994} 
are similar to mosaic plots, but the area of each
tile is proportional to the \emph{expected} cell entry, and each tile
is filled with a number of rectangles corresponding to the observed value.
An \emph{association plot} \cite{vcd:Cohen:1980}
visualizes the standardized deviations of observed frequencies from
those expected under a certain independence hypothesis.
Each cell is represented by a rectangle that has
(signed) height proportional to the residual and width proportional
to the square root of the expected counts, 
so that the area of the box is proportional
to the difference in observed and expected frequencies.

@@

\noindent Current implementations of mosaic displays can be found,
e.g., for \proglang{SAS} \cite{vcd:SAS:2005}, \pkg{ViSta} \cite{vcd:young:1996}, 
\pkg{MANET} \cite{vcd:Unwin+Hawkins+Hofmann:1996},
\pkg{Mondrian} \cite{vcd:Theus:2003},
\proglang{R} \cite{vcd:R:2005}, and \proglang{S-PLUS} \cite{vcd:SPLUS:2005}. 
Implementations of association and sieve plots can only be found in
\proglang{R} and \proglang{SAS} (in the latter, these plots are available
for two-way tables only).
Table~\ref{tab:compare} gives an overview of the available functionality in these systems.
The figures in this chapter have all been produced using the
\proglang{R} system, using the extension packages \pkg{vcd} 
\cite{vcd:Meyer+Zeileis+Hornik:2005} and
\pkg{scatterplot3d} \cite{vcd:ligges+maechler:2003}
(Figure~\ref{fig:bars3d} only), all freely available from the
Comprehensive \proglang{R} Archive Network
(\url{http://CRAN.R-project.org/}). The \proglang{R} code used for the
figures is available from \url{http://statmath.wu-wien.ac.at/projects/vcd/}.

@@

Throughout this section, our examples will be based on
the \data{hospital} data \cite{vcd:wing:1962} given in Table~\ref{tab:hospital}.

@@

Mosaic displays have been introduced by 
\cite{vcd:Hartigan+Kleiner:1981,vcd:Hartigan+Kleiner:1984} and
extended, e.g., by \cite{vcd:Friendly:1994,vcd:Friendly:1999,vcd:Friendly:2000}. 
They visualize the observed values of a contingency table by
area-proportional tiles, arranged in a rectangular mosaic. The tiles are
obtained by recursive partitioning splits of a rectangle.
In the following, we describe the main idea of mosaic plots, 
Chapter III-12 in this book provides more detailed information. Consider our
example of the hospital data from above. Step 1 consists of splitting a square
according to the marginals of one of the variables. To be consistent with the textual
representation, we choose LOS with vertical splits (see Figure~\ref{fig:mosaic1fig}). 
The result is similar to a bar plot where not the height, but the width is
adapted to visualize the counts for each level, which is also 
called a \emph{spine plot} \cite{vcd:hummel:1996}. From this plot, we see that 
the number of patients 
decreases with the length of stay. Step 2 now is to add further splits in the
other direction, i.e., horizontal splits, for the second
variable. This means that each vertical bar is split according to the
marginals of the second variable, \emph{given} the first variable (see
Figure~\ref{fig:mosaic2fig}). The resulting plot visualizes the 
contingency table where each cell has a size proportional to the
corresponding table entry.
We can still see the marginal distribution of LOS, %'
and additionally, the visit frequency \emph{given} each category of LOS. %'
If the two variables were independent, the grid would be regular. Clearly, compared
to a length of stay of 10--19 years, more patients get regular visits for stays from 2--9 
years, and conversely, less patients get regular visits for stays for more than 19 years.
For patients that get no visits, the pattern is reversed.

@@

When we try to explain data, we assume the validity of a certain
model for the generating process. In the case of two-way contingency
tables, the two most common and well-known hypotheses \cite{vcd:Agresti:2002} are

@@

In the last section, we have seen how to compare observed and expected
values of a contingency table using sieve plots. We can do this
more straightforwardly by using a plot that directly visualizes the residuals. The
most widely used residuals are the Pearson residuals

@@

\noindent 
Due to the three-dimensional nature of human color
perception \cite{vcd:Mollon:1995}, 
it has been common to specify colors using the three primaries
red, green, and blue (RGB colors), especially for computer devices. 
The appearance of the on-screen colors is affected by the device
characteristics: For example, the intensity $I$ of a primary 
on a particular device follows the rule $I=L^\gamma$, with $L$
the value for the primary color, and $\gamma$ device-dependent 
(but typically close to 2.2). Therefore, a first caveat is that if colors are to
appear identical on different devices, their gamma characteristics
have to be taken into account.

Choosing colors and color palettes in RGB space can be a bit inconvenient
and hence software implementations are often based on
Hue-Saturation-Value (HSV) colors (or the comparable Hue-Luminance-Satu\-ration 
color scheme). Both spaces are rather similar transformations of the RGB space
\cite{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
colors in many computer packages \cite{vcd:Moretti+Lyons:2002}. Each color in HSV space is
represented by three dimensions $(H, S, V)$: the hue $H$ (dominant wavelength in the spectrum, in $[0, 360]$), 
the saturation $S$ (`colorfulness' or `pureness', in $[0, 100]$), and 
the value $V$ (`brightness', amount of gray, in $[0, 100]$).\footnote{In many implementations,
all three dimensions are scaled to the unit interval instead of the coordinates used here.}
These intuitive dimensions
make HSV colors easyier to specifiy than, e.g, RGB colors. 
However, HSV colors have several disadvantages.
Most importantly, HSV colors are not perceptually
uniform because the three HSV dimensions map only poorly to the three
perceptual dimensions of the human visual system  \cite{vcd:Brewer:1999,vcd:Ihaka:2003}.
One important issue here is that HSV dimensions are confounded, e.g., saturation
is not uniform across different hues.
As an example, see Figure~\ref{fig:colors} (left) 
showing a qualitative color palette (colors $(H, 100, 100)$ for varying hues $H$) in the HSV
space: although saturation and value are fixed, the fully saturated blue is
perceived much darker than the fully saturated red or green,
making it difficult to judge the size of shaded areas.
Furthermore, the flashy fully saturated HSV colors are hard 
to look at for a longer time. For similar reasons, it is equally difficult
to derive acceptable diverging palettes from the HSV space, i.e.,
bipolar scales containing colors ranging between two very distinct colors.
The upper part of Figure~\ref{fig:diverging} shows a diverging
palette in the HSV space with colors ranging from a saturated red $(0, 100, 100)$
over a neutral white $(H, 0, 100)$ to a saturated blue $(240, 100, 100)$.
Although the palette should be balanced with
respect to colorfulness and brightness, the red colors are perceived
to be more more intense and flashy than the corresponding blue
colors.

@@

All plots introduced in Section~\ref{sec:twoway} are composed of empty
tiles. It seems intuitive to use filled tiles to highlight information
of interest. Consider again the hospital example: in Figure~\ref{fig:highlight},
we mark tiles for patients never or seldom visited to visualize their proportion in
the LOS strata. For optical clarity, we set the
spacing between the visit frequency tiles to 0. Clearly, the
proportion of these patients increases with LOS. In fact, such a
``stacked'' plot can also be interpreted as spine plot with highlighting
\cite{vcd:hummel:1996}, particularly useful in analyzing relationships among
categorical data with a binary dependent variable. More variations on
this theme, such as doubledecker plots, are treated in Chapter III-12.

@@

Mosaic plots in their initial version are monochrome
displays. Friendly \cite{vcd:Friendly:1994} introduced a
residual-based shading of the tiles to additionally visualize the
residuals from a given independence model fitted to the table. 
The idea is to use a color coding for the mosaic tiles that visualizes the
sign and absolute size of each residual $r_{ij}$: Cells corresponding to small
residuals ($|r_{ij}| < 2$) have no color. Cells with medium sized residuals
($2 \le |r_{ij}| < 4$) are shaded light blue and light red for positive and negative
residuals, respectively. Cells with large residuals ($|r_{ij}| \ge 4$) are shaded
with a fully saturated blue and red, respectively. 
The heuristic for choosing the cut-offs $2$ and $4$ is that the Pearson residuals
are asymptotically standard normal which implies that the highlighted cells are those with
residuals \emph{individually} significant at approximately the $\alpha
= 0.05$ and $\alpha = 0.0001$ levels. However, the main purpose of the
Friendly shading is not to visualize significance but the \emph{pattern} of
deviation from independence \cite{vcd:Friendly:2000}. 
In addition to the shading of the rectangles themselves, the Friendly shading also
encompasses a choice of line type and line color of the rectangle borders
with similar ideas as described above, useful when no color can be used.

@@

\noindent known to have a limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
An important reason for using the unconditional limiting distribution
for the $X^2$ statistic from (\ref{eq:ChiSquare}) was the closed form result
for the distribution. Recently, with the improving perfomance of computers,
conditional inference (or permutation tests)---carried out either by simulation
or by computation of the (asymptotic) permutation distribution---have been receiving
increasing attention \cite{vcd:Ernst:2004,vcd:Pesarin:2001,vcd:Strasser+Weber:1999}. 
For testing the independence hypothesis from (\ref{eq:H0}),
using a permutation test is particularly intuitive due to the 
permutation invariance (given row and column sums) of this problem.
Consequently, all results in this paper are based on conditional inference performed
by simulating the permutation distribution of test statistics of type
$\lambda([r_{ij}])$.

@@

The heuristic for choosing the cut-off points 
in the Friendly shading may lead to wrong conclusions: especially in
large tables, the test of
independence may not be significant even though some of the residuals are
``large''. On the other hand, the test might be significant even
though the residuals are ``small''. In fact, the cut-off points are really
data-dependent. Consider the case of the \data{arthritis} data 
\cite{vcd:Koch+Edwards:1988}, resulting from a double-blind clinical trial
investigating a new treatment for rheumatoid arthritis, stratified by
gender (see Table~\ref{tab:arthritis} for the female patients):
\begin{table}[h]
\begin{center}
\begin{tabular}{|l|rrr|r|}
\hline
 & \multicolumn{4}{c|}{\textbf{Improvement}}\\
\textbf{Treatment} & None & Some & Marked & $\Sigma$ \\
\hline
Placebo & 19 & 7 & 6 & 32 \\
Treatment & 6 & 5 & 16 & 27 \\\hline
$\Sigma$ & 25 & 12 & 22 & 59 \\
\hline
\end{tabular}
\caption{The \data{arthritis} data (female patients).}
\label{tab:arthritis}
\end{center}
\end{table}
Figure~\ref{fig:arthritis} visualizes the results for the female
patients, again by means of a mosaic display. Clearly, the hypothesis
of independence is rejected by the $\chi^2$ test even on a $1\%$ level 
($p=\Sexpr{round(art_chisq$p.value,4)}$), but
since all residuals are in the 
$[\Sexpr{round(min(chisq.test(art)$residuals),4)},\Sexpr{round(max(chisq.test(art)$residuals),4)}]$ %$
interval, the tiles remain uncolored. A solution to this issue is to use a different
test statistic, for example the maximum of the absolute values of the
Pearson residuals \cite{vcd:Meyer+Zeileis+Hornik:2003} instead of the sum of squares:

@@

\noindent Given a critical value $c_\alpha$ for this test statistic, all residuals whose
absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at level
$\alpha$ \cite[ch. 7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells giving
evidence for the rejection of the independence hypothesis can easily be identified.
As explained above, 
the conditional distribution of this test statistic under the null can be obtained
by simulation, sampling tables with the same row and column sums $n_{i+}$ and $n_{+j}$
using, e.g., the Patefield algorithm \cite{vcd:Patefield:1981} 
and computing the maximum statistic for each of these tables.
In Figure~\ref{fig:arthritis2}, we visualize again the \data{arthritis} data, this
time using the maximum test statistic and its 10\% and 
1\% critical values as cut-off points.
The shading of the
tiles now clearly shows that the treatment is effective: significantly more patients
in the treatment group exhibit marked improvement than would be
expected under independence.

@@

As an example, consider the well-known UCB admissions data
\cite{vcd:Bickel+Hammel+O'Connell:1975} on applicants, classified by
admission and gender, to graduate school at the University of California
in Berkeley for the six largest departments in 1973. The `flattened'
contingency table, putting department into the columns and
admission---nested in gender---into the rows, is given by
Table~\ref{tab:ucb}. The table aggregated over all departments gives
wrong evidence of gender bias, i.e., higher admission rates for male
students.

@@

A first step in the exploratory analysis of more complex tables is to
get a quick overview on the data. For this, 
all basic plots can be combined in pairwise displays,
arranged in a matrix similar to scatterplots in a pairs plot.
The diagonal cells contain the variable names, optionally with
univariate statistics, whereas the off-diagonal cells feature plots whose
variables are implicitly specified by the cells' position in the matrix. %'
In Figure~\ref{fig:pairs}, the diagonal cells show bar plots for the
distribution of the variables, and the off-diagonal cells mosaic plots
for the corresponding pair of variables. The plots suggest that
admission differs between male and female students, and between the departments,
and that the proportion of male and female students varies across the departments.
In particular, departments A and B have a higher proportion of male students
and a lower rejection rate compared to the other departments.

@@

The next step is to investigate three-way interactions among the variables.
In this example, we have a binary variable of interest (admission)
that shall be ``explained'' by the others. 
A natural way of representing such a three-way table is to use a mosaic
display, 
splitting first in the explanatory variables department and gender and highlighting
the resulting mosaic with respect to the dependent variable admissions. Here, we
use the vertical splits for both explanatory variables, resulting in the 
doubledecker plot in Figure~\ref{fig:doubleucb}. From the widths of the
tiles it can be seen that students apply to the six departments in
unequal numbers (with a particularly small
number of females in departments~A and B),
and from the highlighting we see again that the admission rates differ
among the departments (roughly speaking, the admission rate is high for A and B,
low for F and in between for C to E). The rates 
are equal for male and female students, except for department A where
\textit{more} female than male students are admitted.

@@

In addition to the exploratory approaches 
to the visualization of multi-way tables outlined in 
the previous section, there are specialized displays 
designed for the visualization of conditional independence structures.

@@

% Since mosaic displays are conditional plots by construction, we can use them
% for stratified data by using the conditioning variables as first splitting variables. 
% The mosaic in Fig.~\ref{fig:ucbmosaic} shows the distribution of admission
% and gender for all departments. A larger spacing for
% the department variable has been used to simplify the distinction between the 
% groups. 
As we have seen, we can use all plots for the analysis of stratified
data by  first using the conditioning variables for splitting.
However, this also displays the marginal distributions of the conditioning variables
which might make it more difficult to interpret the dependent variables of
interest. Especially for
very unevenly distributed marginals, the strata can become very
distorted, and the tiles, along with their shading, inscrutable.
An alternative is to complement the conditioning in the model by
a conditioning in the plot, i.e., to use trellis-like layouts for
visualizing \emph{partial} tables as defined by the conditioning variables. 
All subtables are corrected for the marginals and thus, all
corresponding plots in the panels have the same size. 
In addition, trellis layouts help to reduce the complexity of bigger tables. 
Figure~\ref{fig:ucbadmissions} visualizes the data by means of a conditional 
association plot. Each panel corresponds to a department, and contains
an association plot corresponding to the partial (fourfold) table of admission
and gender. Accordingly, the shading visualizes the
residuals from the corresponding conditional independence model (independence of gender
and admission, given department), stratified by department. Clearly,
the situation in department A (more women/less men admitted than
expected under the null hypothesis) causes the rejection of the
hypothesis of conditional independence. In Figure~\ref{fig:ucbadmissions},
data-driven cut-offs (derived from the maximum test for conditional independence)
are used, resulting in a more colorful shading than the fixed cut-offs in
Figure~\ref{fig:ucbmosaic} and \ref{fig:ucbassoc}.

@@

>>>>

\usepackage{Sweave,hyperref,color,graphicx,makeidx,multicol} 
\usepackage[bottom]{footmisc}% places footnotes at page bottom

@@

%\makeindex

@@

\index{Categorical data} Categorical data analysis is typically based on two- or
higher-dimensional \index{contingency tables} contingency tables, cross-tabulating the
co-occurences of levels of nominal and/or ordinal data. 
In order to explain these, statisticians
typically look for (conditional) independence structures using common
methods such as independence tests and
log-linear models. One idea of visualization techniques is to
use the human visual system to detect structures in the data that
possibly are not obvious from solely numeric output (e.g., test
statistics). Whether the
task is purely exploratory or model based, techniques such as mosaic, sieve, and
association plots offer good support for visualization. Especially
mosaic and sieve plots have been extended over the last two decades, 
and implementations exist in many statistical environments.

All three graphical methods visualize aspects of (possibly higher-dimensional)
contingency tables. A \emph{mosaic plot} \index{mosaic plot}
\cite{vcd:Hartigan+Kleiner:1984} is basically 
an area-proportional visualization of (typically, observed) frequencies, composed
of tiles (corresponding to the cells) created by recursive
vertical and horizontal splits of a rectangle. Thus, the area of each tile
is proportional to the corresponding cell entry \emph{given} the
dimensions of previous splits. 
\emph{Sieve plots} \index{sieve plot} \cite{vcd:riedwyl+schuepbach:1994} 
are similar to mosaic plots, but the area of each
tile is proportional to the \emph{expected} cell entry, and each tile
is filled with a number of rectangles corresponding to the observed value.
An \emph{association plot} \index{association plot} \cite{vcd:Cohen:1980}
visualizes the standardized deviations of observed frequencies from
those expected under a certain independence hypothesis.
Each cell is represented by a rectangle that has
(signed) height proportional to the residual and width proportional
to the square root of the expected counts, 
so that the area of the box is proportional
to the difference in observed and expected frequencies.

@@

\noindent Current implementations of mosaic displays can be found,
e.g., for \proglang{SAS} \cite{vcd:SAS:2005}, \pkg{ViSta} \cite{vcd:young:1996}, 
\pkg{MANET} \cite{vcd:Unwin+Hawkins+Hofmann:1996},
\pkg{Mondrian} \cite{vcd:Theus:2003},
\proglang{R} \cite{vcd:R:2005}, and \proglang{S-PLUS} \cite{vcd:SPLUS:2005}. 
Implementations of association and sieve plots can only be found in
\proglang{R} and \proglang{SAS} (in the latter, these plots are available
for two-way tables only).
Table~\ref{tab:compare} gives an overview of the available functionality in these systems.
The figures in this chapter have all been produced using the
\proglang{R} \index{R} system, using the extension packages \pkg{vcd} 
\cite{vcd:Meyer+Zeileis+Hornik:2005} and
\pkg{scatterplot3d} \cite{vcd:ligges+maechler:2003}
(Figure~\ref{fig:bars3d} only), all freely available from the
Comprehensive \proglang{R} Archive Network
(\url{http://CRAN.R-project.org/}). The \proglang{R} code used for the
figures is available from \url{http://statmath.wu-wien.ac.at/projects/vcd/}.

@@

Throughout this section, our examples will be based on
the \index{Data!hospital} 
\data{hospital} data \cite{vcd:wing:1962} given in Table~\ref{tab:hospital}.

@@

\index{mosaic plot} Mosaic displays have been introduced by 
\cite{vcd:Hartigan+Kleiner:1981,vcd:Hartigan+Kleiner:1984} and
extended, e.g., by \cite{vcd:Friendly:1994,vcd:Friendly:1999,vcd:Friendly:2000}. 
They visualize the observed values of a contingency table by
area-proportional tiles, arranged in a rectangular mosaic. The tiles are
obtained by recursive partitioning splits of a rectangle.
In the following, we describe the main idea of mosaic plots, 
Chapter III-12 in this book provides more detailed information. Consider our
example of the hospital data from above. Step 1 consists of splitting a square
according to the marginals of one of the variables. To be consistent with the textual
representation, we choose LOS with vertical splits (see Figure~\ref{fig:mosaic1fig}). 
The result is similar to a bar plot where not the height, but the width is
adapted to visualize the counts for each level, which is also 
called a \emph{spine plot} \cite{vcd:hummel:1996}. \index{spine plot}
From this plot, we see that the number of patients 
decreases with the length of stay. Step 2 now is to add further splits in the
other direction, i.e., horizontal splits, for the second
variable. This means that each vertical bar is split according to the
marginals of the second variable, \emph{given} the first variable (see
Figure~\ref{fig:mosaic2fig}). The resulting plot visualizes the 
contingency table where each cell has a size proportional to the
corresponding table entry.
We can still see the marginal distribution of LOS, %'
and additionally, the visit frequency \emph{given} each category of LOS. %'
If the two variables were independent, the grid would be regular. Clearly, compared
to a length of stay of 10--19 years, more patients get regular visits for stays from 2--9 
years, and conversely, less patients get regular visits for stays for more than 19 years.
For patients that get no visits, the pattern is reversed.

@@

\index{sieve plot}
When we try to explain data, we assume the validity of a certain
model for the generating process. In the case of two-way contingency
tables, the two most common and well-known hypotheses \cite{vcd:Agresti:2002} are

@@

\index{association plot}
In the last section, we have seen how to compare observed and expected
values of a contingency table using sieve plots. We can do this
more straightforwardly by using a plot that directly visualizes the residuals. The
most widely used residuals are the Pearson residuals \index{Pearson residuals}

@@

\noindent 
Due to the three-dimensional nature of human color
perception \cite{vcd:Mollon:1995}, 
it has been common to specify colors using the three primaries
red, green, and blue (RGB colors), especially for computer devices. 
\index{RGB colors}
The appearance of the on-screen colors is affected by the device
characteristics: For example, the intensity $I$ of a primary 
on a particular device follows the rule $I=L^\gamma$, with $L$
the value for the primary color, and $\gamma$ device-dependent 
(but typically close to 2.2). Therefore, a first caveat is that if colors are to
appear identical on different devices, their gamma characteristics
have to be taken into account.

Choosing colors and color palettes in RGB space can be a bit inconvenient
and hence software implementations are often based on \index{HSV colors}
Hue-Saturation-Value (HSV) colors (or the comparable Hue-Luminance-Satu\-ration 
color scheme). Both spaces are rather similar transformations of the RGB space
\cite{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
colors in many computer packages \cite{vcd:Moretti+Lyons:2002}. Each color in HSV space is
represented by three dimensions $(H, S, V)$: the hue $H$ (dominant wavelength in the spectrum, in $[0, 360]$), 
the saturation $S$ (`colorfulness' or `pureness', in $[0, 100]$), and 
the value $V$ (`brightness', amount of gray, in $[0, 100]$).\footnote{In many implementations,
all three dimensions are scaled to the unit interval instead of the coordinates used here.}
These intuitive dimensions
make HSV colors easyier to specifiy than, e.g, RGB colors. 
However, HSV colors have several disadvantages.
Most importantly, HSV colors are not perceptually
uniform because the three HSV dimensions map only poorly to the three
perceptual dimensions of the human visual system  \cite{vcd:Brewer:1999,vcd:Ihaka:2003}.
One important issue here is that HSV dimensions are confounded, e.g., saturation
is not uniform across different hues.
As an example, see Figure~\ref{fig:colors} (left) 
showing a qualitative color palette \index{qualitative color palette}
(colors $(H, 100, 100)$ for varying hues $H$) in the HSV
space: although saturation and value are fixed, the fully saturated blue is
perceived much darker than the fully saturated red or green,
making it difficult to judge the size of shaded areas.
Furthermore, the flashy fully saturated HSV colors are hard 
to look at for a longer time. For similar reasons, it is equally difficult
to derive acceptable diverging palettes from the HSV space, i.e.,
\index{diverging color palette}
bipolar scales containing colors ranging between two very distinct colors.
The upper part of Figure~\ref{fig:diverging} shows a diverging
palette in the HSV space with colors ranging from a saturated red $(0, 100, 100)$
over a neutral white $(H, 0, 100)$ to a saturated blue $(240, 100, 100)$.
Although the palette should be balanced with
respect to colorfulness and brightness, the red colors are perceived
to be more more intense and flashy than the corresponding blue
colors.

@@

All plots introduced in Section~\ref{sec:twoway} are composed of empty
tiles. It seems intuitive to use filled tiles to highlight information
of interest. Consider again the hospital example: in Figure~\ref{fig:highlight},
we mark tiles for patients never or seldom visited to visualize their proportion in
the LOS strata. For optical clarity, we set the
spacing between the visit frequency tiles to 0. Clearly, the
proportion of these patients increases with LOS. In fact, such a
``stacked'' plot can also be interpreted as spine plot with
highlighting \index{highlighting}
\cite{vcd:hummel:1996}, particularly useful in analyzing relationships among
categorical data with a binary dependent variable. More variations on
this theme, such as doubledecker plots, are treated in Chapter III-12.

@@

Mosaic plots in their initial version are monochrome
displays. Friendly \cite{vcd:Friendly:1994} introduced a
\index{residual-based shading} residual-based shading
of the tiles to additionally visualize the
residuals from a given independence model fitted to the table. 
The idea is to use a color coding for the mosaic tiles that visualizes the
sign and absolute size of each residual $r_{ij}$: Cells corresponding to small
residuals ($|r_{ij}| < 2$) have no color. Cells with medium sized residuals
($2 \le |r_{ij}| < 4$) are shaded light blue and light red for positive and negative
residuals, respectively. Cells with large residuals ($|r_{ij}| \ge 4$) are shaded
with a fully saturated blue and red, respectively. 
The heuristic for choosing the cut-offs $2$ and $4$ is that the Pearson residuals
are asymptotically standard normal which implies that the highlighted cells are those with
residuals \emph{individually} significant at approximately the $\alpha
= 0.05$ and $\alpha = 0.0001$ levels. However, the main purpose of the
Friendly shading is not to visualize significance but the \emph{pattern} of
deviation from independence \cite{vcd:Friendly:2000}. 
In addition to the shading of the rectangles themselves, the Friendly shading also
encompasses a choice of line type and line color of the rectangle borders
with similar ideas as described above, useful when no color can be used.

@@

\noindent known to have a limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
An important reason for using the unconditional limiting distribution
for the $X^2$ statistic from (\ref{eq:ChiSquare}) was the closed form result
for the distribution. Recently, with the improving perfomance of computers,
conditional inference (or permutation tests)---carried out either by simulation
or by computation of the (asymptotic) permutation distribution---have been receiving
increasing attention \cite{vcd:Ernst:2004,vcd:Pesarin:2001,vcd:Strasser+Weber:1999}. 
For testing the independence hypothesis from (\ref{eq:H0}),
using a \index{permutation test} permutation test is particularly intuitive due to the 
permutation invariance (given row and column sums) of this problem.
Consequently, all results in this paper are based on conditional inference performed
by simulating the permutation distribution of test statistics of type
$\lambda([r_{ij}])$.

@@

The heuristic for choosing the cut-off points 
in the Friendly shading may lead to wrong conclusions: especially in
large tables, the test of
independence may not be significant even though some of the residuals are
``large''. On the other hand, the test might be significant even
though the residuals are ``small''. In fact, the cut-off points are really
data-dependent. Consider the case of the \data{arthritis} data \index{Data!arthritis}
\cite{vcd:Koch+Edwards:1988}, resulting from a double-blind clinical trial
investigating a new treatment for rheumatoid arthritis, stratified by
gender (see Table~\ref{tab:arthritis} for the female patients):
\begin{table}[h]
\begin{center}
\begin{tabular}{|l|rrr|r|}
\hline
 & \multicolumn{4}{c|}{\textbf{Improvement}}\\
\textbf{Treatment} & None & Some & Marked & $\Sigma$ \\
\hline
Placebo & 19 & 7 & 6 & 32 \\
Treatment & 6 & 5 & 16 & 27 \\\hline
$\Sigma$ & 25 & 12 & 22 & 59 \\
\hline
\end{tabular}
\caption{The \data{arthritis} data (female patients).}
\label{tab:arthritis}
\end{center}
\end{table}
Figure~\ref{fig:arthritis} visualizes the results for the female
patients, again by means of a mosaic display. Clearly, the hypothesis
of independence is rejected by the $\chi^2$ test even on a $1\%$ level 
($p=\Sexpr{round(art_chisq$p.value,4)}$), but
since all residuals are in the 
$[\Sexpr{round(min(chisq.test(art)$residuals),4)},\Sexpr{round(max(chisq.test(art)$residuals),4)}]$ %$
interval, the tiles remain uncolored. A solution to this issue is to use a different
test statistic, for example the maximum of the absolute values of the
Pearson residuals \cite{vcd:Meyer+Zeileis+Hornik:2003} instead of the sum of squares:

@@

\noindent Given a critical value $c_\alpha$ for this test statistic, all residuals whose
absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at level
$\alpha$ \cite[ch. 7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells giving
evidence for the rejection of the independence hypothesis can easily be identified.
As explained above, 
the conditional distribution of this test statistic under the null can be obtained
by simulation, sampling tables with the same row and column sums $n_{i+}$ and $n_{+j}$
using, e.g., the Patefield algorithm \cite{vcd:Patefield:1981}
\index{Patefield algorithm}
and computing the maximum statistic for each of these tables.
In Figure~\ref{fig:arthritis2}, we visualize again the \data{arthritis} data, this
time using the maximum test statistic and its 10\% and 
1\% critical values as cut-off points.
The shading of the
tiles now clearly shows that the treatment is effective: significantly more patients
in the treatment group exhibit marked improvement than would be
expected under independence.

@@

As an example, consider the well-known UCB admissions data
\index{Data!UCB Admissions}
\cite{vcd:Bickel+Hammel+O'Connell:1975} on applicants, classified by
admission and gender, to graduate school at the University of California
in Berkeley for the six largest departments in 1973. The `flattened'
contingency table, putting department into the columns and
admission---nested in gender---into the rows, is given by
Table~\ref{tab:ucb}. The table aggregated over all departments gives
wrong evidence of gender bias, i.e., higher admission rates for male
students.

@@

A first step in the exploratory analysis of more complex tables is to
get a quick overview on the data. For this, 
all basic plots can be combined in pairwise displays,
arranged in a matrix similar to scatterplots in a pairs
plot. \index{pairs plot}
The diagonal cells contain the variable names, optionally with
univariate statistics, whereas the off-diagonal cells feature plots whose
variables are implicitly specified by the cells' position in the matrix. %'
In Figure~\ref{fig:pairs}, the diagonal cells show bar plots for the
distribution of the variables, and the off-diagonal cells mosaic plots
for the corresponding pair of variables. The plots suggest that
admission differs between male and female students, and between the departments,
and that the proportion of male and female students varies across the departments.
In particular, departments A and B have a higher proportion of male students
and a lower rejection rate compared to the other departments.

@@

The next step is to investigate three-way interactions among the variables.
In this example, we have a binary variable of interest (admission)
that shall be ``explained'' by the others. 
A natural way of representing such a three-way table is to use a mosaic
display, 
splitting first in the explanatory variables department and gender and highlighting
the resulting mosaic with respect to the dependent variable admissions. Here, we
use the vertical splits for both explanatory variables, resulting in the 
\index{doubledecker plot} doubledecker plot 
in Figure~\ref{fig:doubleucb}. From the widths of the
tiles it can be seen that students apply to the six departments in
unequal numbers (with a particularly small
number of females in departments~A and B),
and from the highlighting we see again that the admission rates differ
among the departments (roughly speaking, the admission rate is high for A and B,
low for F and in between for C to E). The rates 
are equal for male and female students, except for department A where
\textit{more} female than male students are admitted.

@@

In addition to the exploratory approaches 
to the visualization of multi-way tables outlined in 
the previous section, there are specialized displays 
designed for the visualization of \index{conditional independence} 
conditional independence structures.

@@

% Since mosaic displays are conditional plots by construction, we can use them
% for stratified data by using the conditioning variables as first splitting variables. 
% The mosaic in Fig.~\ref{fig:ucbmosaic} shows the distribution of admission
% and gender for all departments. A larger spacing for
% the department variable has been used to simplify the distinction between the 
% groups. 
As we have seen, we can use all plots for the analysis of stratified
data by  first using the conditioning variables for splitting.
However, this also displays the marginal distributions of the conditioning variables
which might make it more difficult to interpret the dependent variables of
interest. Especially for
very unevenly distributed marginals, the strata can become very
distorted, and the tiles, along with their shading, inscrutable.
An alternative is to complement the conditioning in the model by
a conditioning in the plot, i.e., to use trellis-like layouts for
\index{trellis layout}
visualizing \emph{partial} tables as defined by the conditioning variables. 
All subtables are corrected for the marginals and thus, all
corresponding plots in the panels have the same size. 
In addition, trellis layouts help to reduce the complexity of bigger tables. 
Figure~\ref{fig:ucbadmissions} visualizes the data by means of a conditional 
association plot. Each panel corresponds to a department, and contains
an association plot corresponding to the partial (fourfold) table of admission
and gender. Accordingly, the shading visualizes the
residuals from the corresponding conditional independence model (independence of gender
and admission, given department), stratified by department. Clearly,
the situation in department A (more women/less men admitted than
expected under the null hypothesis) causes the rejection of the
hypothesis of conditional independence. In Figure~\ref{fig:ucbadmissions},
data-driven cut-offs (derived from the maximum test for conditional independence)
are used, resulting in a more colorful shading than the fixed cut-offs in
Figure~\ref{fig:ucbmosaic} and \ref{fig:ucbassoc}.

@@

\printindex
